%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgments}{acknowledgments}

\begin{flushright}{\slshape    
    We have seen that computer programming is an art, \\ 
    because it applies accumulated knowledge to the world, \\ 
    because it requires skill and ingenuity, and especially \\
    because it produces objects of beauty.} \\ \medskip
    --- \defcitealias{knuth:1974}{Donald E. Knuth}\citetalias{knuth:1974} \citep{knuth:1974}
\end{flushright}



\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgments}
I would first like to thank my thesis advisor Andreas Simbürger of the Fakultät für Informatik at Passau Universität. He was always there to help me and provide me guidances whenever I needed it. 
I also want to take Michael Krüse of Inria laboratory in Paris, my mentor during the Google Summer of Code who provide me very helpful support.

\bigskip

I would also like to thank all my teachers since the beginning of my cursus who gave me the love of computer science.

\bigskip

Thank you to Laurence and all my friends who support me during this long effort.
Thank you to all others students involved in the double diplom Insa Lyon - Universität Passau.

\bigskip

I also want to express my appreciation for my girlfriend Adeline who stuck with me during the long months of working with my computer. This accomplishment would not have been possible without you. 

\bigskip

Finalement, je voudrais exprimer ma plus profonde gratitude à mes parents, Murielle et Didier, à mes sœurs Séverine et Elisabeth, à mon beau-frère Marc et à toute ma famille pour leur soutien sans faille durant toute ma scolarité. Sans eux, rien n'aurait été possible. Merci beaucoup !

\endgroup



